---
title: "Using doseRider for studying Non-Linear Dose-Response in BPA."
author: "Pablo Monfort-Lanzas"
date: "2023-08-01"
output: html_document
---

# Using doseRider for studying Non-Linear Dose-Response in BPA

The `doseRider` package is used for analyzing dose-response relationships in gene expression data using Generalized Additive Mixed Models (GAMMs). This report focuses on bisphenol A (BPA) and its alternatives, analyzing their impact on gene expression in MCF-7 cells.

## RMarkdown Workflow for doseRider Analysis

### 1. Loading required packages

```{r, warning=FALSE, echo=FALSE, include=FALSE}
library(doseRider)
library(DESeq2)
library(ggplot2)
library(dplyr)
```

### 2. Data Import

Data from PRJNA869442 (GSE211183) has been preprocessed for this analysis.

```{r}
load("../data/PRJNA869442.rda")
```

#### Dose Selection for Each Compound

Doses were selected based on criteria like precipitation, stress response biomarkers, and cell proliferation.

```{r}
compound_doses <- list(
  "BPA" = c(0.0005, 0.1, 0.5, 1, 5, 10)
)
  doses <- c(0,compound_doses["BPA"][[1]])
  
# Example filtering for BPA
bpa_data <- PRJNA869442[,(colData(PRJNA869442)$Chemical %in% c("Cells, no treatment","BPA")) &
                             (colData(PRJNA869442)$Dose %in% doses)]
bpa_data$sample <- colnames(bpa_data)
```

### 3. Differential Expression Analysis

```{r}
colData(bpa_data)$Dose <- unlist(colData(bpa_data)$Dose)
#colData(bpa_data)$Dose <- log(unlist(colData(bpa_data)$Dose) + 1)
dds <- DESeqDataSetFromMatrix(countData = assay(bpa_data),
                              colData   = colData(bpa_data),
                              design    = ~ Dose*Dose)

dds <- DESeq(dds, parallel = TRUE, quiet = TRUE)
res <- results(dds)

filter_res <- as.data.frame(res[(res$baseMean > 11) & (res$pvalue < 0.5),])
bpa_data <- estimate_model_parameters(bpa_data)

# Filter low express genes
filter_bpa_data <- bpa_data[rownames(bpa_data) %in% rownames(filter_res),]
```

### 4. Load Gene Sets

```{r}
gmt <- loadCPDB("Symbol")
top_geneset <- read.csv("../../Projects/TOX/ToxicoDataset/results/TG-GATES/WeightedScoreCPDBTGGATES.tsv", sep = ",")
gmt_id <- doseRider:::filter_gmt_by_id(gmt, unique(top_geneset[abs(top_geneset$WeightedScore) > 2,]$X))
```

### 5. Dose-Response Model with doseRider

```{r}
dose_rider_results <- DoseRider(se = filter_bpa_data, gmt = gmt_id, 
                                           dose_col = "Dose", omic = "rnaseq", 
                                           minGSsize = 50, maxGSsize = 1000, 
                                           method = "fdr", covariates = c(),
                                          modelType = "LMM")

res_df <- as.data.frame.DoseRider(dose_rider_results)
lKeep <- (res_df$Adjusted_non_linear_P_Value < 0.01) & (res_df$Adjusted_Linear_P_Value < 0.01)
table((res_df$Adjusted_non_linear_P_Value < 0.01) & (res_df$Adjusted_Linear_P_Value < 0.01))
table(res_df$Best_Model_AICc)
#save(res, file = "../../doseRider_save/res_bpa_20_1000.rda")
```

## Results and Discussion

```{r}
top <- 15
save_path <- "/home/monfortl/doseRider/plots/"

p1 <- dose_response_heatmap(dose_rider_results, dose_col = "Dose", top = top, order_column = "Adjusted_non_linear_P_Value", decreasing = F )
plotFile1 <- paste0(save_path,"/plot1.jpeg")
p2 <- plot_gene_set_random_effects(dose_rider_results, dose_col = "Dose", top = top)
plotFile2 <- paste0(save_path,"/plot2.jpeg")
p3 <- plot_top_pathway_responses(dose_rider_results, top = 9, ncol = 3)
plotFile3 <- paste0(save_path,"/plot3.jpeg")
p4 <- plot_gene_random_effect_relationship(dose_rider_results, "Androgen receptor signaling pathway")
plotFile4 <- paste0(save_path,"/plot4.jpeg")
p5 <- plot_dotplot_top_pathways(dose_rider_results, top = top, order_column = "Genes", decreasing = T)
plotFile5 <- paste0(save_path,"/plot5.jpeg")
p6 <- create_gene_heatmap(dose_rider_results,dose_col = "Dose", gene_set_name = "Androgen receptor signaling pathway")
plotFile6 <- paste0(save_path,"/plot6.jpeg")

jpeg(file=plotFile1,width = 600, height = 500, units = "px")
 plot(p1)
dev.off()
ggsave(plotFile2, plot = p2, width = 250, height = 150, units = "mm", dpi = 600)
ggsave(plotFile3, plot = p3, width = 400, height = 200, units = "mm", dpi = 600)
ggsave(plotFile4, plot = p4, width = 400, height = 200, units = "mm", dpi = 600)
ggsave(plotFile5, plot = p5, width = 400, height = 200, units = "mm", dpi = 600)
jpeg(file=plotFile6,width = 700, height = 700, units = "px")
 plot(p6)
dev.off()
```

# Microarray example:

dataset1: Human Transcriptomic Integrated Approach for Testing and Assessment Oral Reference Dose for Steatosis: 0.00, 0.3125, 0.625, 1.25, 2.50, 5.00 (24 samples) GSE135117

dataset2: Primary hepatocyte cultures from 3 human donors were treated with TCDD and PCB 126 compounds, with 14 samples per compound (6 doses) (6.5 to 14 log10 M doses for TCDD and 5 to 12 log10 M doses for PCB 126). GSE14553

dataset3: The cell used in this study is THP-1 macrophages from Homo sapiens. They were exposed to rigid multi-walled carbon nanotubes (rCNT) at three concentrations (5, 10, or 20 µg/mL) and untreated controls at 24, 48, and 72 hours, with each group performed as triplicates. GSE146708


The cell used is BEAS-2B bronchial epithelial cells (Homo sapiens), exposed to N-hydroxylated PhIP at nine different doses ranging from 1 x 10-6 to 5 x 10-10 M, with triplicate cell cultures for each dose. There is no mention of donors in this dataset. PRJNA151269; GEO: GSE34635

